---
title: "kate_analysis"
author: "Kate Colvin"
date: "2026-04-26"
output: pdf_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(readr)
library(gtsummary)
library(gt)
library(gridExtra)
library(broom)

```

Examine the relationship between potential risk factors and intraoperative blood loss, with particular attention to whether these associations differ across race/ethnicity subgroups.
- Previous studies have found that age and sex are associated with intraoperative blood loss; be sure to account for them in your analysis.

```{r}

data <- read_csv("ebl.csv") %>% 
  mutate(sex = factor(sex, levels = c(0, 1),
                      labels = c("Male", "Female")),
         race = factor(race, levels = c(0, 1, 2, 3),
                       labels = c("White", "Black", "Asian", "Hispanic")), 
         op = factor(op, levels = c(0, 1), 
                     labels = c("Body/Tail Resection", "Whipple")), 
         ven_res = factor(ven_res, levels = c(0, 1), 
                          labels = c("No", "Yes")), 
         neo_chemo = factor(neo_chemo, levels = c(0, 1), 
                            labels = c("No", "Yes")),
         neo_xrt = factor(neo_xrt, levels = c(0, 1), 
                          labels = c("No", "Yes")), 
         prot_energy = factor(prot_energy, levels = c(0, 1), 
                              labels = c("No", "Yes")))

# Adding NAs as "Missing"
data_unknown <- data %>% 
  mutate(race = ifelse(is.na(race), "Unknown", as.character(race)), 
         race = factor(race)) %>% 
  mutate(race = factor(race, levels = c("White", "Black", "Asian", "Hispanic", "Unknown")))

```



## Exploratory Data Analysis

```{r}

table1_unknown <- data_unknown %>%
  select(-c(patid)) %>%
  tbl_summary(by = race,
              label = list(age ~ "Age (years)",
                           sex ~ "Sex", 
                           liver_test ~ "Bilirubin (mg/dL)",
                           op ~ "Operation Type",
                           ebl ~ "Estimated Blood Loss (mL)",
                           ven_res ~ "Vascular Resection",
                           neo_chemo ~ "Neoadjuvant Chemotherapy",
                           neo_xrt ~ "Neoadjuvant Radiation",
                           les_size ~ "Tumor Size (cm)",
                           or_time ~ "Operative Time (minutes)", 
                           prot_energy ~ "Adequate Protein-Energy")) %>% 
  add_overall() %>%
  bold_labels() %>% 
  add_p() %>% 
  bold_p() %>% 
  modify_caption("**Table 1**: Summary of Patient Characteristics by Race") 

table1_unknown

table1_gt <- table1 %>% as_gt()
gtsave(table1_gt, "results/table1.png")

```


```{r}

# Checking distribution of ebl

ggplot(data, aes(x = ebl)) +
  geom_histogram(bins = 30, fill = "navy") +
  labs(title = "Distribution of Estimated Blood Loss", x = "EBL (mL)") +
  theme_minimal()

ggplot(data, aes(sample = ebl)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(
    title = "QQ Plot of Estimated Blood Loss (EBL)",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles") +
  theme_minimal()

```

```{r}

# EBL by race

# Histogram 
ggplot(data_unknown, aes(x = race, y = ebl, color = race)) +
  geom_boxplot() +
  labs(x = "Race/Ethnicity", y = "EBL (mL)") + 
  theme_minimal()

# Ridge plot
data_unknown %>% ggplot(aes(x = ebl, y = race, fill = race)) +
   ggridges::geom_density_ridges(
     alpha = 0.85,
     scale = 1.2,
     color = "white") +
   theme_minimal() +
   labs(x = "EBL (mL)", y = "Race",
     title = "Distribution of Estimated Blood Loss by Race") +
   theme(legend.position = "none", 
         plot.title = element_text(size = 12))

```

```{r}

# Checking if there are differences in covariate x ebl distribution by race

# Categorical Variables

p1 <- data_unknown %>% filter(race != "Unknown") %>% 
  ggplot(aes(x = race, y = ebl, fill = factor(ven_res))) +
  geom_boxplot() +
  labs(title = "EBL by Race and Vascular Resection",
       fill = "Ven. Resection") +
  theme_minimal() +
  theme(plot.title = element_text(size = 11), 
        legend.key.size = unit(0.25, "cm"), 
        legend.title = element_text(size = 9))
p1


p2 <- data_unknown %>% filter(race != "Unknown") %>% 
  ggplot(aes(x = race, y = ebl, fill = factor(op))) +
  geom_boxplot() +
  labs(title = "EBL by Race and Operation Type",
       fill = "Operation") +
  theme_minimal() +
  theme(plot.title = element_text(size = 11), 
        legend.key.size = unit(0.1, "cm"), 
        legend.title = element_text(size = 9))
p2

p3 <- data_unknown %>% filter(race != "Unknown") %>% 
  ggplot(aes(x = race, y = ebl, fill = factor(neo_chemo))) +
  geom_boxplot() +
  labs(title = "EBL by Race and Neoadjuvant Chemotherapy",
       fill = "Chemo") +
  theme_minimal() +
  theme(plot.title = element_text(size = 11), 
        legend.key.size = unit(0.25, "cm"), 
        legend.title = element_text(size = 9))
p3

p4 <- data_unknown %>% filter(race != "Unknown") %>% 
  ggplot(aes(x = race, y = ebl, fill = factor(neo_xrt))) +
  geom_boxplot() +
  labs(title = "EBL by Race and Neoadjuvant Radiation",
       fill = "Radiation") +
  theme_minimal() +
  theme(plot.title = element_text(size = 11), 
        legend.key.size = unit(0.25, "cm"), 
        legend.title = element_text(size = 9))
p4

cat_figs <- grid.arrange(p1, p2, p3, p4, ncol = 2)
cat_figs


ggsave(filename = "results/cat_figs.png", plot = cat_figs, 
       width = 9, height = 5, dpi = 300)

```


```{r}

# Continuous Variables

p5 <- data_unknown %>% filter(race != "Unknown") %>% 
  ggplot(aes(x = or_time, y = ebl, fill = race, color = race)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess") +
  facet_wrap(~ race) +
  labs(title = "EBL vs Operative Time by Race") +
  theme_minimal() +
  theme(plot.title = element_text(size = 11))
p5

p6 <- data_unknown %>% filter(race != "Unknown") %>% 
  ggplot(aes(x = les_size, y = ebl, fill = race, color = race)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess") +
  facet_wrap(~ race) +
  labs(title = "EBL vs Tumor Size by Race") +
  theme_minimal() +
  theme(legend.position = "none", 
         plot.title = element_text(size = 11))
p6

p7 <- data_unknown %>% filter(race != "Unknown") %>% 
  ggplot(aes(x = liver_test, y = ebl, fill = race, color = race)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ race) +
  labs(title = "EBL vs Bilirubin by Race") +
  theme_minimal() +
  theme(legend.position = "none", 
         plot.title = element_text(size = 11))
p7

p8 <- data_unknown %>% filter(race != "Unknown") %>% 
  ggplot(aes(x = age, y = ebl, fill = race, color = race)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ race) +
  labs(title = "EBL vs Age by Race") +
  theme_minimal() +
  theme(legend.position = "none", 
         plot.title = element_text(size = 11))
p8


cont_figs <- grid.arrange(p8, p5, p6, p7, ncol = 2)
cont_figs


ggsave(filename = "results/cont_figs.png", plot = cont_figs, 
       width = 9, height = 5, dpi = 300)

```

- Categorical Variables: 
  - Bottom row; not enough non-white people to include interaction term
  - Top left; not enough non-white people, could combine the groups into white vs. non-white (probably won't)
- Continuous Variables: Slopes are all very similar, or not enough people


## Primary Modeling, complete case analysis

Which interaction terms to include, based on figures above:
- Categorical: ven_res (INSIGNIFICANT, DROPPED IT), op
- Continuous: or_time (steeper slopes?) (INSIGNIFICANT, DROPPED IT), liver_test (SMALL SAMPLES, REMOVED), les_size (ASIAN ONLY SIGNIFICANT, PROBABLY USNTABLE)

For now, just including op x race. 

```{r}

model <- lm(ebl ~ age + sex + race + liver_test + op + ven_res + 
              neo_chemo + neo_xrt + les_size + or_time + 
              op * race, 
            data = data)
# summary(model)


model_results <- tbl_regression(
  model,
  label = list(age ~ "Age",
               sex ~ "Sex", 
               race ~ "Race", 
               liver_test ~ "Bilirubin (mg/dL)",
               op ~ "Operation Type", 
               ven_res ~ "Vascular Resection",
               neo_chemo ~ "Neoadjuvant Chemotherapy",
               neo_xrt ~ "Neoadjuvant Radiation",
               les_size ~ "Tumor Size (cm)",
               or_time ~ "Operative Time (minutes)"), 
  conf.int = TRUE,
  conf.level = 0.95) %>%
  bold_labels() %>% 
  bold_p() %>% 
  modify_caption("**Table X**: Primary Model Results")

model_results

model_results_gt <- model_results %>% as_gt()
gtsave(model_results_gt, "results/table_model_results.png")

```

```{r}

# Model Diagnostics 

par(mfrow = c(2,2))
plot(model)

```


```{r}

# LRT, checking that the inclusion of the interaction term helped

model_main <- lm(ebl ~ age + sex + race + liver_test + op + ven_res + 
              neo_chemo + neo_xrt + les_size + or_time, 
            data = data)

anova(model_main, model)

# Significant!

```



## Checking Race Missingness, Sensitivity Analysis

Checking possible mechanisms of race missingness (missing 59 races)

```{r}

data %>%
  mutate(race_missing = is.na(race)) %>%
  select(-c(patid, race)) %>%
  tbl_summary(by = race_missing, 
              label = list(age ~ "Age (years)",
                           sex ~ "Sex", 
                           liver_test ~ "Bilirubin (mg/dL)",
                           op ~ "Operation Type",
                           ebl ~ "Estimated Blood Loss (mL)",
                           ven_res ~ "Vascular Resection",
                           neo_chemo ~ "Neoadjuvant Chemotherapy",
                           neo_xrt ~ "Neoadjuvant Radiation",
                           les_size ~ "Tumor Size (cm)",
                           or_time ~ "Operative Time (minutes)", 
                           prot_energy ~ "Adequate Protein-Energy")) %>%
  modify_caption("**Table X**: Summary of Patient Characteristics by Race Missingness") %>% 
  add_p()

```


```{r}

na_model <- glm(is.na(race) ~ age + sex + ebl + op + ven_res +neo_chemo + neo_xrt + les_size + or_time,
                data = data,
                family = binomial)

summary(na_model)

# ven_resYes, neo_chemoYes statistically significantly associated with missing race

```


```{r}

library(mice)

imputed <- data %>% select(-c(patid)) %>% 
  mice(m = 20, method = "pmm", seed = 123)

fit_imputed <- with(imputed,
  lm(ebl ~ age + sex + race + liver_test + op + ven_res +
       neo_chemo + neo_xrt + les_size + or_time +
       op * race))

pooled <- pool(fit_imputed)
summary(pooled)

```

```{r}

# Imputation results 

pooled_tidy <- tidy(pooled, conf.int = TRUE)

pooled_clean <- pooled_tidy %>%
  select(term, estimate, conf.low, conf.high, p.value)

tbl_mi <- pooled_clean %>%
  gt() %>%
  cols_label(
    term = "Variable",
    estimate = "Estimate",
    conf.low = "Lower CI",
    conf.high = "Upper CI",
    p.value = "p-value") %>%
  fmt_number(columns = c(estimate, conf.low, conf.high), decimals = 2) %>%
  fmt_number(columns = p.value, decimals = 3) %>%
  tab_header(title = "Model Results (Multiple Imputation)")
tbl_mi

```


```{r}

complete_data <- complete(imputed, "long")

race_dist <- complete_data %>%
  group_by(.imp, race) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(race) %>%
  summarise(mean_n = mean(n))
race_dist

```


