---
title: "kate_analysis"
author: "Kate Colvin"
date: "2026-04-26"
output: pdf_document
---

- Unknown race model (bonus)
- Update EDA with unknown

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(readr)
library(gtsummary)
library(gt)
library(gridExtra)
library(broom)

```

Examine the relationship between potential risk factors and intraoperative blood loss, with particular attention to whether these associations differ across race/ethnicity subgroups.
- Previous studies have found that age and sex are associated with intraoperative blood loss; be sure to account for them in your analysis.

```{r}

data <- read_csv("ebl.csv") %>% 
  mutate(sex = factor(sex, levels = c(0, 1),
                      labels = c("Male", "Female")),
        # race = factor(race, levels = c(0, 1, 2, 3),
        #               labels = c("White", "Black", "Asian", "Hispanic")), 
         op = factor(op, levels = c(0, 1), 
                     labels = c("Body/Tail Resection", "Whipple")), 
         ven_res = factor(ven_res, levels = c(0, 1), 
                          labels = c("No", "Yes")), 
         neo_chemo = factor(neo_chemo, levels = c(0, 1), 
                            labels = c("No", "Yes")),
         neo_xrt = factor(neo_xrt, levels = c(0, 1), 
                          labels = c("No", "Yes")), 
         prot_energy = factor(prot_energy, levels = c(0, 1), 
                              labels = c("No", "Yes")))

# Race Binary
data <- data %>% mutate(
    race_bin = case_when(race == 0 ~ "White", 
                         race %in% c(1, 2, 3) ~ "Non-White",
                         TRUE ~ NA_character_),
    race_bin = factor(race_bin, 
                      levels = c("White", "Non-White")))

# Adding NAs as "Missing"
data_unknown <- data %>% 
  mutate(race = factor(race, levels = c(0, 1, 2, 3),
                labels = c("White", "Black", "Asian", "Hispanic"))) %>%  
  mutate(race_bin = ifelse(is.na(race_bin), "Unknown", as.character(race_bin)), 
         race_bin = factor(race_bin)) %>% 
  mutate(race_bin = factor(race_bin, levels = c("White", "Non-White", "Unknown"))) %>% 
  mutate(race = ifelse(is.na(race), "Unknown", as.character(race)), 
         race = factor(race)) %>% 
  mutate(race = factor(race, levels = c("White", "Black", "Asian", "Hispanic", "Unknown")))

data <- data %>% mutate(race = factor(race, levels = c(0, 1, 2, 3),
                        labels = c("White", "Black", "Asian", "Hispanic")))

```



## Exploratory Data Analysis

```{r}

table1_unknown <- data_unknown %>%
  select(-c(patid, race_bin)) %>%
  tbl_summary(by = race,
              label = list(age ~ "Age (years)",
                           sex ~ "Sex", 
                           liver_test ~ "Bilirubin (mg/dL)",
                           op ~ "Operation Type",
                           ebl ~ "Estimated Blood Loss (mL)",
                           ven_res ~ "Vascular Resection",
                           neo_chemo ~ "Neoadjuvant Chemotherapy",
                           neo_xrt ~ "Neoadjuvant Radiation",
                           les_size ~ "Tumor Size (cm)",
                           or_time ~ "Operative Time (minutes)", 
                           prot_energy ~ "Adequate Protein-Energy")) %>% 
  add_overall() %>%
  bold_labels() %>% 
  add_p() %>% 
  bold_p() %>% 
  modify_caption("**Table 1**: Summary of Patient Characteristics by Race") 

table1_unknown

table1_gt <- table1_unknown %>% as_gt()
gtsave(table1_gt, "results/table1.png")

```


```{r}

# Checking distribution of ebl

ebl_dist <- ggplot(data, aes(x = ebl)) +
  geom_histogram(bins = 30, fill = "navy") +
  labs(title = "Distribution of Estimated Blood Loss", x = "EBL (mL)") +
  theme_minimal()
ebl_dist

ggplot(data, aes(sample = ebl)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(
    title = "QQ Plot of Estimated Blood Loss (EBL)",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles") +
  theme_minimal()

```

```{r}

# EBL by race

# Histogram 
ggplot(data_unknown, aes(x = race_bin, y = ebl, color = race_bin)) +
  geom_boxplot() +
  labs(x = "Race/Ethnicity", y = "EBL (mL)") + 
  theme_minimal()

# Ridge plot
ebl_ridge <- data_unknown %>% ggplot(aes(x = ebl, y = race_bin, fill = race_bin)) +
   ggridges::geom_density_ridges(
     alpha = 0.85,
     scale = 1.2,
     color = "white") +
   theme_minimal() +
   labs(x = "EBL (mL)", y = "Race",
     title = "Distribution of Estimated Blood Loss by Race") +
   theme(legend.position = "none", 
         plot.title = element_text(size = 12))
ebl_ridge

ebl_dists <- grid.arrange(ebl_dist, ebl_ridge, ncol = 1)
ebl_dists

ggsave(filename = "results/ebl_dists.png", plot = ebl_dists, 
       width = 6, height = 5, dpi = 300)


# Ridge plot by distinct racial groups
ebl_ridge_race <- data_unknown %>% ggplot(aes(x = ebl, y = race, fill = race)) +
   ggridges::geom_density_ridges(
     alpha = 0.85,
     scale = 1.2,
     color = "white") +
   theme_minimal() +
   labs(x = "EBL (mL)", y = "Race",
     title = "Distribution of Estimated Blood Loss by Race") +
   theme(legend.position = "none", 
         plot.title = element_text(size = 12))
ebl_ridge_race

```

```{r}

# Checking if there are differences in covariate x ebl distribution by race

# Categorical Variables

p1 <- data_unknown %>% 
  ggplot(aes(x = race_bin, y = ebl, fill = factor(ven_res))) +
  geom_boxplot() +
  labs(title = "EBL by Race and Vascular Res.",
       fill = "Vascular Res.", 
       x = "") +
  theme_minimal() +
  theme(plot.title = element_text(size = 10), 
        legend.key.size = unit(0.25, "cm"), 
        legend.title = element_text(size = 9))
p1


p2 <- data_unknown %>% 
  ggplot(aes(x = race_bin, y = ebl, fill = factor(op))) +
  geom_boxplot() +
  labs(title = "EBL by Race and Operation Type",
       fill = "Operation", 
       x = "") +
  theme_minimal() +
  theme(plot.title = element_text(size = 10), 
        legend.key.size = unit(0.1, "cm"), 
        legend.title = element_text(size = 9))
p2

p3 <- data_unknown %>% 
  ggplot(aes(x = race_bin, y = ebl, fill = factor(neo_chemo))) +
  geom_boxplot() +
  labs(title = "EBL by Race and Neoadjuvant Chemo.",
       fill = "Chemo", 
       x = "") +
  theme_minimal() +
  theme(plot.title = element_text(size = 10), 
        legend.key.size = unit(0.25, "cm"), 
        legend.title = element_text(size = 9))
p3

p4 <- data_unknown %>% 
  ggplot(aes(x = race_bin, y = ebl, fill = factor(neo_xrt))) +
  geom_boxplot() +
  labs(title = "EBL by Race and Neoadjuvant Rad.",
       fill = "Radiation", 
       x = "") +
  theme_minimal() +
  theme(plot.title = element_text(size = 10), 
        legend.key.size = unit(0.25, "cm"), 
        legend.title = element_text(size = 9))
p4

p9 <- data_unknown %>% 
  ggplot(aes(x = race_bin, y = ebl, fill = factor(sex))) +
  geom_boxplot() +
  labs(title = "EBL by Race and Sex",
       fill = "Radiation", 
       x = "") +
  theme_minimal() +
  theme(plot.title = element_text(size = 10), 
        legend.key.size = unit(0.25, "cm"), 
        legend.title = element_text(size = 9))
p9


cat_figs <- grid.arrange(p1, p2, p9, p3, p4, ncol = 3)
cat_figs

ggsave(filename = "results/cat_figs.png", plot = cat_figs, 
       width = 12, height = 5, dpi = 300)

```


```{r}

# Continuous Variables

p5 <- data_unknown %>% 
  ggplot(aes(x = or_time, y = ebl, fill = race_bin, color = race_bin)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess") +
  facet_wrap(~ race_bin) +
  labs(title = "EBL vs Operative Time by Race", 
       x = "Operative Time", 
       color = "Race", fill = "Race") +
  theme_minimal() +
  theme(plot.title = element_text(size = 11))
p5

p6 <- data_unknown %>% 
  ggplot(aes(x = les_size, y = ebl, fill = race_bin, color = race_bin)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess") +
  facet_wrap(~ race_bin) +
  labs(title = "EBL vs Tumor Size by Race", 
       x = "Tumor Size (cm)") +
  theme_minimal() +
  theme(legend.position = "none", 
         plot.title = element_text(size = 11))
p6

p7 <- data_unknown %>% 
  ggplot(aes(x = liver_test, y = ebl, fill = race_bin, color = race_bin)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ race_bin) +
  labs(title = "EBL vs Bilirubin by Race", 
       x = "Bilirubin (mg/DL)") +
  theme_minimal() +
  theme(legend.position = "none", 
         plot.title = element_text(size = 11))
p7

p8 <- data_unknown %>% 
  ggplot(aes(x = age, y = ebl, fill = race_bin, color = race_bin)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ race_bin) +
  labs(title = "EBL vs Age by Race", 
       x = "Age") +
  theme_minimal() +
  theme(legend.position = "none", 
         plot.title = element_text(size = 11))
p8


cont_figs <- grid.arrange(p8, p5, p6, p7, ncol = 2)
cont_figs


ggsave(filename = "results/cont_figs.png", plot = cont_figs, 
       width = 12, height = 5, dpi = 300)

```



## Primary Modeling, complete case analysis

Which interaction terms to include, based on figures above:
- Categorical: Operation type
- Continuous: Operative Time

$$
\text{EBL}_i \sim N(\mu_i, \sigma^2)
$$

$$
\begin{aligned}
\text{EBL}_i =\;& \beta_0 
+ \beta_1 \text{Age}_i 
+ \beta_2 \mathbf{1}(\text{Female}_i)
+ \beta_3 \mathbf{1}(\text{Non-White}_i)
+ \beta_4 \text{Bilirubin}_i 
+ \beta_5 \mathbf{1}(\text{Whipple}_i) 
+ \beta_6 \mathbf{1}(\text{Vascular Resection}_i) \\
&+ \beta_7 \mathbf{1}(\text{Neoadj. Chemo}_i) 
+ \beta_8 \mathbf{1}(\text{Neoadj. Radiation}_i) 
+ \beta_9 \text{Tumor Size}_i 
+ \beta_{10} \text{Operative Time}_i \\
&+ \beta_{11} \left[\mathbf{1}(\text{Whipple}_i) \times \mathbf{1}(\text{Non-White}_i)\right] 
+ \beta_{12} \left[\text{Operative Time}_i \times \mathbf{1}(\text{Non-White}_i)\right]
+ \varepsilon_i
\end{aligned}
$$


```{r}

model <- lm(ebl ~ age + sex + race_bin + liver_test + op + ven_res + 
              neo_chemo + neo_xrt + les_size + or_time + 
              op * race_bin + or_time * race_bin, 
            data = data)
summary(model)


model_results <- tbl_regression(
  model,
  label = list(age ~ "Age (years)",
               sex ~ "Sex", 
               race_bin ~ "Race", 
               liver_test ~ "Bilirubin (mg/dL)",
               op ~ "Operation Type", 
               ven_res ~ "Vascular Resection",
               neo_chemo ~ "Neoadjuvant Chemotherapy",
               neo_xrt ~ "Neoadjuvant Radiation",
               les_size ~ "Tumor Size (cm)",
               or_time ~ "Operative Time (minutes)"), 
  conf.int = TRUE,
  conf.level = 0.95) %>%
  bold_labels() %>% 
  bold_p() %>% 
  modify_caption("**Table 2**: Model Results (Complete-Case)")

model_results

model_results_gt <- model_results %>% as_gt()
gtsave(model_results_gt, "results/table_model_results.png")

```


```{r}

# F-test, checking that the inclusion of the interaction terms helped

model_main <- lm(ebl ~ age + sex + race_bin + liver_test + op + ven_res + 
              neo_chemo + neo_xrt + les_size + or_time, 
              data = data)

ft <- anova(model_main, model)

# Not significant!

ft_tbl <- as.data.frame(ft) %>%
  tibble::rownames_to_column(var = "Model") %>% 
  select(Model, everything()) %>% 
  rename(
    df = Df,
    RSS = RSS,
    `p-value` = `Pr(>F)`) %>% 
  mutate(`p-value` = signif(`p-value`, 3)) %>% 
  select(Model, df, RSS, `p-value`)

ft_tbl$Model <- c("Reduced (no interaction)", "Full (with interaction)")
ft_tbl

gt(ft_tbl) %>% 
tab_header(title = "Table X: F-Test, Inclusion of Interaction Terms") %>% 
gtsave("results/table_ft.png")

```

```{r}

# No-interaction model as main model

model_results_noint <- tbl_regression(
  model_main,
  label = list(age ~ "Age (years)",
               sex ~ "Sex", 
               race_bin ~ "Race", 
               liver_test ~ "Bilirubin (mg/dL)",
               op ~ "Operation Type", 
               ven_res ~ "Vascular Resection",
               neo_chemo ~ "Neoadjuvant Chemotherapy",
               neo_xrt ~ "Neoadjuvant Radiation",
               les_size ~ "Tumor Size (cm)",
               or_time ~ "Operative Time (minutes)"), 
  conf.int = TRUE,
  conf.level = 0.95) %>%
  bold_labels() %>% 
  bold_p() %>% 
  modify_caption("**Table 3**: Model Results, No Interactions (Complete-Case)")

model_results_noint

model_results_noint_gt <- model_results_noint %>% as_gt()
gtsave(model_results_noint_gt, "results/table_model_results_noint.png")

```


```{r}

# Model Diagnostics 

png("results/model_diagnostics.png", width = 1200, height = 800)

par(mfrow = c(2,2))
plot(model_main)

dev.off()

```


## Checking Race Missingness, Sensitivity Analysis

Checking possible mechanisms of race missingness (missing 59 races)

```{r}

data %>%
  mutate(race_missing = is.na(race)) %>%
  select(-c(patid, race)) %>%
  tbl_summary(by = race_missing, 
              label = list(age ~ "Age (years)",
                           sex ~ "Sex", 
                           liver_test ~ "Bilirubin (mg/dL)",
                           op ~ "Operation Type",
                           ebl ~ "Estimated Blood Loss (mL)",
                           ven_res ~ "Vascular Resection",
                           neo_chemo ~ "Neoadjuvant Chemotherapy",
                           neo_xrt ~ "Neoadjuvant Radiation",
                           les_size ~ "Tumor Size (cm)",
                           or_time ~ "Operative Time (minutes)", 
                           prot_energy ~ "Adequate Protein-Energy")) %>%
  modify_caption("**Table X**: Summary of Patient Characteristics by Race Missingness") %>% 
  add_p()

```


```{r}

na_model <- glm(is.na(race) ~ age + sex + liver_test + ebl + op + ven_res + neo_chemo + neo_xrt + les_size + or_time,
                data = data,
                family = binomial)

summary(na_model)

# ven_resYes, neo_chemoYes statistically significantly associated with missing race
# MAR

na_model_table <- tbl_regression(
  na_model,
  label = list(age ~ "Age (years)",
               sex ~ "Sex", 
               liver_test ~ "Bilirubin (mg/dL)",
               ebl ~ "Estimated Blood Loss (mL)", 
               op ~ "Operation Type", 
               ven_res ~ "Vascular Resection",
               neo_chemo ~ "Neoadjuvant Chemotherapy",
               neo_xrt ~ "Neoadjuvant Radiation",
               les_size ~ "Tumor Size (cm)",
               or_time ~ "Operative Time (minutes)"), 
  conf.int = TRUE,
  conf.level = 0.95) %>%
  bold_labels() %>% 
  bold_p() %>% 
  modify_caption("**Table 4**: Race Missingness Model Results")

na_model_table

na_model_table_gt <- na_model_table %>% as_gt()
gtsave(na_model_table_gt, "results/table_na_model.png")


```


```{r}

library(mice)

data_mice <- data %>% select(-c(patid, race_bin, prot_energy))

meth <- make.method(data_mice)
meth["race"] <- "polyreg"

imputed <- data_mice %>% 
  mice(m = 20, method = meth, seed = 1)

fit_imputed <- with(imputed, {
  race_bin <- factor(
    ifelse(race == "White", "White", "Non-White"),
    levels = c("White", "Non-White"))
  lm(ebl ~ age + sex + race_bin + liver_test + op + ven_res + 
       neo_chemo + neo_xrt + les_size + or_time)})

pooled <- pool(fit_imputed)
summary(pooled)

```

```{r}

# Imputation results 

pooled_tidy <- tidy(pooled, conf.int = TRUE)

pooled_clean <- pooled_tidy %>%
  select(term, estimate, conf.low, conf.high, p.value) %>% 
  mutate(term = recode(term,
    "(Intercept)" = "Intercept",
    "age" = "Age (years)",
    "sexFemale" = "Female (vs Male)",
    "race_binNon-White" = "Non-White (vs White)",
    "liver_test" = "Bilirubin (mg/dL)",
    "opWhipple" = "Whipple (vs Body/Tail)",
    "ven_resYes" = "Vascular Resection",
    "neo_chemoYes" = "Neoadjuvant Chemotherapy",
    "neo_xrtYes" = "Neoadjuvant Radiation",
    "les_size" = "Tumor Size (cm)",
    "or_time" = "Operative Time (minutes)")) 

tbl_mi <- pooled_clean %>%
  gt() %>%
  cols_label(
    term = "Variable",
    estimate = "Estimate",
    conf.low = "Lower CI",
    conf.high = "Upper CI",
    p.value = "p-value") %>%
  fmt_number(columns = c(estimate, conf.low, conf.high), decimals = 2) %>%
  fmt_number(columns = p.value, decimals = 3) %>%
  tab_header(title = "Table 5: Model Results, No Interactions (Multiple Imputation)") %>% 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      columns = p.value,
      rows = as.numeric(p.value) < 0.05)) %>% 
  text_transform(
    locations = cells_body(columns = p.value),
    fn = function(x) {
      x <- as.numeric(x)
      ifelse(
        is.na(x), "",
        ifelse(
          x < 0.001, "<0.001",
          sprintf("%.3f", x)))})
tbl_mi

gtsave(tbl_mi, "results/table_mi.png")

```


Additional no interaction model; unknown as it's own category

```{r}

model_unknown <- lm(ebl ~ age + sex + race_bin + liver_test + op + ven_res + 
              neo_chemo + neo_xrt + les_size + or_time + age*race_bin, 
            data = data_unknown)
summary(model_unknown)


model_unknown_results <- tbl_regression(
  model_unknown,
  label = list(age ~ "Age",
               sex ~ "Sex", 
               race_bin ~ "Race", 
               liver_test ~ "Bilirubin (mg/dL)",
               op ~ "Operation Type", 
               ven_res ~ "Vascular Resection",
               neo_chemo ~ "Neoadjuvant Chemotherapy",
               neo_xrt ~ "Neoadjuvant Radiation",
               les_size ~ "Tumor Size (cm)",
               or_time ~ "Operative Time (minutes)"), 
  conf.int = TRUE,
  conf.level = 0.95) %>%
  bold_labels() %>% 
  bold_p() %>% 
  modify_caption("**Table 6**: Model Results Including Unknown Race Patients")

model_unknown_results 

model_unknown_results_gt <- model_unknown_results %>% as_gt()
gtsave(model_unknown_results_gt, "results/table_model_results_unknown.png")


```

