---
title: "kate_analysis"
author: "Kate Colvin"
date: "2026-04-26"
output: pdf_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(readr)
library(gtsummary)
library(gt)

```

Examine the relationship between potential risk factors and intraoperative blood loss, with particular attention to whether these associations differ across race/ethnicity subgroups.
- Previous studies have found that age and sex are associated with intraoperative blood loss; be sure to account for them in your analysis.

```{r}

data <- read_csv("ebl.csv") %>% 
  mutate(sex = factor(sex, levels = c(0, 1),
                      labels = c("Male", "Female")),
         race = factor(race, levels = c(0, 1, 2, 3),
                       labels = c("White", "Black", "Asian", "Hispanic")), 
         op = factor(op, levels = c(0, 1), 
                     labels = c("Body/Tail Resection", "Whipple")), 
         ven_res = factor(ven_res, levels = c(0, 1), 
                          labels = c("No", "Yes")), 
         neo_chemo = factor(neo_chemo, levels = c(0, 1), 
                            labels = c("No", "Yes")),
         neo_xrt = factor(neo_xrt, levels = c(0, 1), 
                          labels = c("No", "Yes")), 
         prot_energy = factor(prot_energy, levels = c(0, 1), 
                              labels = c("No", "Yes")))

# Adding NAs as "Missing"
data_unknown <- data %>% 
  mutate(race = ifelse(is.na(race), "Unknown", as.character(race)), 
         race = factor(race)) %>% 
  mutate(race = factor(race, levels = c("White", "Black", "Asian", "Hispanic", "Unknown")))

```



## Exploratory Data Analysis

```{r}

table1_unknown <- data_unknown %>%
  select(-c(patid)) %>%
  tbl_summary(by = race,
              label = list(age ~ "Age (years)",
                           sex ~ "Sex", 
                           liver_test ~ "Bilirubin (mg/dL)",
                           op ~ "Operation Type",
                           ebl ~ "Estimated Blood Loss (mL)",
                           ven_res ~ "Vascular Resection",
                           neo_chemo ~ "Neoadjuvant Chemotherapy",
                           neo_xrt ~ "Neoadjuvant Radiation",
                           les_size ~ "Tumor Size (cm)",
                           or_time ~ "Operative Time (minutes)", 
                           prot_energy ~ "Adequate Protein-Energy")) %>% 
  add_overall() %>%
  bold_labels() %>% 
  add_p() %>% 
  bold_p() %>% 
  modify_caption("**Table 1**: Summary of Patient Characteristics by Race")

table1_unknown

# table1_gt <- table1 %>% as_gt()
# gtsave(table1_gt, "results/table1.png")

```


```{r}

# Checking distribution of ebl

ggplot(data, aes(x = ebl)) +
  geom_histogram(bins = 30, fill = "navy") +
  labs(title = "Distribution of Estimated Blood Loss", x = "EBL (mL)") +
  theme_minimal()

ggplot(data, aes(sample = ebl)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(
    title = "QQ Plot of Estimated Blood Loss (EBL)",
    x = "Theoretical Quantiles",
    y = "Sample Quantiles") +
  theme_minimal()

```

```{r}

# EBL by race

# Histogram 
ggplot(data_unknown, aes(x = race, y = ebl, color = race)) +
  geom_boxplot() +
  labs(x = "Race/Ethnicity", y = "EBL (mL)") + 
  theme_minimal()

# Ridge plot
data_unknown %>% ggplot(aes(x = ebl, y = race, fill = race)) +
   ggridges::geom_density_ridges(
     alpha = 0.85,
     scale = 1.2,
     color = "white") +
   theme_minimal() +
   labs(x = "EBL (mL)", y = "Race",
     title = "Distribution of Estimated Blood Loss by Race") +
   theme(legend.position = "none", 
         plot.title = element_text(size = 12))

```


```{r}



```



## Checking Race Missingness

Checking possible mechanisms of race missingness (missing 59 races)

```{r}

data %>%
  mutate(race_missing = is.na(race)) %>%
  select(-c(patid, race)) %>%
  tbl_summary(by = race_missing, 
              label = list(age ~ "Age (years)",
                           sex ~ "Sex", 
                           liver_test ~ "Bilirubin (mg/dL)",
                           op ~ "Operation Type",
                           ebl ~ "Estimated Blood Loss (mL)",
                           ven_res ~ "Vascular Resection",
                           neo_chemo ~ "Neoadjuvant Chemotherapy",
                           neo_xrt ~ "Neoadjuvant Radiation",
                           les_size ~ "Tumor Size (cm)",
                           or_time ~ "Operative Time (minutes)", 
                           prot_energy ~ "Adequate Protein-Energy")) %>%
  modify_caption("**Table X**: Summary of Patient Characteristics by Race Missingness") %>% 
  add_p()

```



```{r}

na_model <- glm(is.na(race) ~ age + sex + ebl + op + ven_res +neo_chemo + neo_xrt + les_size + or_time,
                data = data,
                family = binomial)

summary(na_model)

# ven_resYes, neo_chemoYes statistically significantly associated with missing race

```


## Modeling

```{r}



```

